---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

In this notebook, we will do things..
- load some packages
- discuss directories
- read data 
- make a plot using ggplot 


Some package we will use (install these if you haven't already)
```{r}
library(tidyverse)
library(here)
library(ggplot2)
```
# A note on directories

Where does here() point to?
```{r}
here::here()
here::here("data")
```

```{r}
dat <- read_csv(here::here("data", "prob_simu1.csv"))
```
The data is now loaded into the variable `dat`, which is a particular data frame called a tibble. Let's take a look at it.
```{r}
dat
```

manipulation using dplyr 

add a new column using mutate
```{r}
dat %>% 
  mutate(prob_rounded = round(prob, 2))
```

filter the data
```{r}
dat %>% 
  filter(prob > 0.5)
```





# Data visualization

In this training, you will be provided with code to make plots. This section introduces the kind of code that you may see. 


Let's start by making a histogram of our data, how would that work?

```{r}
dat %>% # we start with the tibble 
  ggplot(aes(x = prob)) +
  geom_histogram()
```

