---
title: "Brms - getting started"
author: "Leontine Alkema"
date: "August 21, 2024"
output:
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

# Installation

See instructions in slide set. Once installed, you should be able to load the brms library:

```{r}
library(brms)
```

We will also use the posterior package to summarize the output:

```{r}
library(posterior)
```

# Example 1: use $n$ data points with unknown variance

Step 1: simulate data

```{r}
set.seed(123) # to make sure we get the same data every time
n <- 100 # number of data points
y <- rnorm(n, 0, 1) # generate n data points from a normal distribution with mean 0 and sd 1
hist(y)
```

We'll just store the data in an object called dat

```{r}
dat <- tibble(y = y)
```

Now let's do some model fitting using brm! We use default priors. 
(Note: adding cache = TRUE in R chunk to avoid having to refit every time you knit the document).

```{r,  cache = TRUE}
fit <- brm(y ~ 1, # formula to state we want to estimate the mean 
                  # we'll discuss formula in more detail in the next module
           data = dat, # passing in the data 
           seed = 12345, # making the fit reproducible
           chains = 4, cores = getOption("mc.cores", 4) 
           # 4 chains run in parallel on 4 cores
           )
```

Let's check the output. The standard summary function gives estimates of all parameters and some diagnostics. Can you find the estimate for $\mu$? and its diagnostics? 
```{r}
summary(fit)
```

You can also pull out whatever you're interested in using the summarise_draws function, e.g.
```{r}
draws <- as_draws_array(fit, variable = "Intercept") # this is our mu
summarise_draws(draws, default_summary_measures())
```

To get a 95% CI:
```{r}
summarise_draws(draws, ~quantile(.x, probs = c(0.025, 0.975)))
```

These overview plots show, for each parameter, its histogram with posterior samples and its trace plot.
```{r}
plot(fit, variable = c("Intercept", "sigma"))
```

# Example 2: everything is normal

Let $y|\mu \sim N(\mu, s_y^2)$, with $s_y$ known, and set prior $\mu \sim N(m_0, s_0^2)$.

Here is a function that calls brm to fit the Bayesian model, using the following argument names:

-   y_observed for the observed value y,
-   sd_y for its SD,
-   and mean_muprior, sd_muprior for the prior mean and SD of $\mu$.

```{r}
fit_brm_normalnormal <- function(y_observed, sd_y, mean_muprior, sd_muprior){
  mu_prior <- set_prior(paste0("normal(", mean_muprior, ",", sd_muprior, ")"), class = "Intercept")
  sigma_fixed <- set_prior(paste0("constant(", sd_y, ")"), class = "sigma")
  fit  <- brm(y ~ 1, family = gaussian(), data = tibble(y = y_observed), 
              cores = getOption("mc.cores", 4),
              seed = 12345, 
              prior = c(mu_prior, sigma_fixed)
              )
  return(fit)
}
```

Let's fit the model using the function we just defined:
```{r, cache = TRUE}
fit2  <- fit_brm_normalnormal(y_observed = 0.3, sd_y = 0.05, mean_muprior = 0.5, sd_muprior = 0.1)
```

Note that we fixed sigma so it's not estimated. 
```{r}
summary(fit2)
```

```{r}
plot(fit2, variable = c("Intercept"))
```

Are these results the same as when using the closed-form expression?
```{r}
draws <- as_draws_array(fit2, variable = "Intercept") # this is our mu
summarise_draws(draws, default_summary_measures())
```
To get a 95% CI:
```{r}
summarise_draws(draws, ~quantile(.x, probs = c(0.025, 0.975)))
```
The End!
