---
title: "Bayesian hierarchical modeling - exercise 2 V2"
author: "Leontine Alkema"
date: "September 24, 2024"
output:
  pdf_document:
    number_sections: true
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(brms)
library(posterior)
library(tidybayes)
```

In this notebook, we fit a hierarchical model to FP data to estimate country-year-specific mCPR using data after 2010.
Note that this is a simplified exercise, for example, we do not take account of sampling errors and we assume that mCPR changes linearly on the logit-scale. It's just a start to get familiar with basic hierarchical-temporal models.

# Read and process FP data


```{r}
dat0 <- read_csv(here::here("data/fpet_track20data2024.csv"), col_types = list(division_numeric_code = "c"))
region_info <- read_csv(here::here("data/fpet_regions_all.csv"), col_types = list(division_numeric_code = "c"))
```


```{r}
dat1 <- dat0 %>%
  filter(start_date >= 2010) %>% # use obs after 2010
  filter(is_in_union == "Y") %>% # married women! 
  rename(mcpr = contraceptive_use_modern) %>%
  mutate(year = 1/2*(start_date + end_date)) %>% # define survey year as midpoint 
  filter(!is.na(mcpr), mcpr >= 0.005) %>% # for keeping things simple, remove missing values and mcpr less than 0.5% (convenience)
  select(division_numeric_code, mcpr, year) %>%
  left_join(region_info) %>% # add in country name and region info
  select(division_numeric_code, mcpr, 
        year,
         name_country, name_region, name_sub_region) %>% # UNPD regions  
  filter(name_region == "Africa") # here we focus just on countries in Africa
dat1
```

Let's estimate mcpr on the logit-scale. We define transformation functions to make this easier.

```{r}
logit <- function(p) log(p/(1-p))
inv_logit <- function(x) exp(x)/(1+exp(x))
```

We add y = logit(mcpr) to our data set and rename the division_numeric_code to iso for convenience.
```{r}
dat <- 
  dat1 %>%
  mutate(y = logit(mcpr)) %>%
  rename(iso = division_numeric_code) %>%  # 'cause 3 letters are better than many
  mutate(yearmin2016 = year - 2016) # we also add variable yearmin2016 which is year - 2016
dat
```

Create summary data set with info for each country:
```{r}
set.seed(12345) # used for creating jittered nobs
datcountry <- dat %>%
  group_by(iso) %>%
  summarize(nobs = n(), ybar = mean(y), iso = iso[1], country = name_country[1],
            name_region = name_region[1], name_sub_region = name_sub_region[1]) %>%
  mutate(nobs_jitter = nobs*exp(runif (length(nobs), -.1, .1)))
ncountries <- dim(datcountry)[1]
datcountry
```


# Model fitting 

Fit0 is just estimating country mean levels over the period since 2010
```{r,  cache = TRUE}
fit0 <- brm(y ~ (1|iso), 
        data = dat, 
        iter = 4000, warmup = 1000, 
        seed = 1234, 
      cores = getOption("mc.cores", 4))
```

```{r}
summary(fit0)
```


We estimate country-year-specific prevalence using a hierarchical model with country-specific means and shared slopes. 
```{r,  cache = TRUE}
fit <- brm(y ~ (1|iso) + yearmin2016, 
        data = dat, 
        iter = 4000, # adding some more iterations to avoid low effective sample size per chain
        warmup = 1000,
        chains = 4,
         # this is an MCMC tuning parameter, to be discussed later 
       # control = list(adapt_delta = 0.95), 
        seed = 1234, 
      cores = getOption("mc.cores", 4))
```

Summary of model fit: 
```{r}
summary(fit)
```

Make rates of change country-specific and add an extra level in the hierarchy

```{r,  cache = TRUE}
fit2 <- brm(y ~ 1 + yearmin2016 + # added in the year predictor in V2
              (1 + yearmin2016|iso) +  (1 + yearmin2016|name_sub_region)  + yearmin2016,
        data = dat, 
        iter = 2000, 
        seed = 123, 
        # this is an MCMC tuning parameter, to be discussed later 
        control = list(adapt_delta = 0.99, max_treedepth = 14), 
        chains = 4,
        cores = getOption("mc.cores", 4))
```

```{r}
summary(fit2)
```

# Explore output


Select a country and years of interest
```{r}
namec <- "Ethiopia"
predyears <- 2010:2025 # years of interest
```

To produce the estimates, we just need a data frame with the same variables as used in the formula (ie predictors of outcome mu_ct). 
Easiest to take one row from the countrydat and add all years of interest as yearmin2016 
```{r}
preddat <- tibble(
  datcountry %>% filter(country == namec),
  yearmin2016 = predyears - 2016)
preddat
```

We can use the posterior_epred function to create estimates of mcpr over time for a specific country. 
This function just uses the function and wrangles the output into a more convenient format. 
We also save 3 random posterior samples to be able to add those to the plot. 

```{r}
get_est <- function(fit, preddat, predyears) {
  nsamples <- 3
  mcpr_st <- inv_logit(posterior_epred(fit, newdata = preddat))
  set.seed(123456) # for obtaining just a few random posterior samples
  samp_index <- sample(1:nrow(mcpr_st), nsamples)
  colnames(mcpr_st) <- predyears
  est <- bind_cols(mcpr_st, tibble(sample = 1:nrow(mcpr_st))) %>%
    pivot_longer(cols = -sample, names_to = "year", values_to = "value") %>%
      mutate(year = as.numeric(year)) %>%
    group_by(year) %>%
    reframe(low = quantile(value, 0.025), 
              median = quantile(value, 0.5), 
              up = quantile(value, 0.975),
            # some hard coding here, probably a better way!
              samp1 = value[samp_index[1]],
                        samp2 = value[samp_index[2]],
  
                        samp3 = value[samp_index[3]]
  
            ) 
  return(est)
}
```


get estimates based on the 3 models 
```{r}
est0 <- get_est(fit0, preddat, predyears)
est <- get_est(fit, preddat, predyears)
est2 <- get_est(fit2, preddat, predyears)
```


Example output
```{r}
est 
```

Plot 
```{r}
dat %>%
  filter(name_country == namec) %>%
  ggplot(aes(x = year, y = mcpr)) +
  ggplot2::geom_ribbon(data = est, 
    aes(ymin = `low`, y = median, ymax = `up`), fill  = "blue", alpha = 0.1) +
 geom_line(data = est, aes(x = year, y = median), color = "blue") +
 geom_line(data = est, aes(x = year, y = samp1), color = "darkgrey") +
 geom_line(data = est, aes(x = year, y = samp2), color = "darkgrey") +
 geom_line(data = est, aes(x = year, y = samp3), color = "darkgrey") +
  theme_minimal() +
   geom_point() +
  labs(title = namec) + # increase font sizes
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 14))
```

Plot  for model 2
```{r}
dat %>%
  filter(name_country == namec) %>%
  ggplot(aes(x = year, y = mcpr)) +
  ggplot2::geom_ribbon(data = est2, 
    aes(ymin = `low`, y = median, ymax = `up`), fill  = "red", alpha = 0.1) +
 geom_line(data = est2, aes(x = year, y = median), color = "red") +
 geom_line(data = est2, aes(x = year, y = samp1), color = "darkgrey") +
 geom_line(data = est2, aes(x = year, y = samp2), color = "darkgrey") +
 geom_line(data = est2, aes(x = year, y = samp3), color = "darkgrey") +
  theme_minimal() +
   geom_point() +
  labs(title = namec) + # increase font sizes
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 14))
```

Plot  for model 0
```{r}
dat %>%
  filter(name_country == namec) %>%
  ggplot(aes(x = year, y = mcpr)) +
  ggplot2::geom_ribbon(data = est0, 
    aes(ymin = `low`, y = median, ymax = `up`), fill  = "blue", alpha = 0.1) +
 geom_line(data = est0, aes(x = year, y = median), color = "blue") +
 geom_line(data = est0, aes(x = year, y = samp1), color = "darkgrey") +
 geom_line(data = est0, aes(x = year, y = samp2), color = "darkgrey") +
 geom_line(data = est0, aes(x = year, y = samp3), color = "darkgrey") +
  theme_minimal() +
   geom_point() +
  labs(title = namec) + # increase font sizes
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 14))
```

Plot with two fits

```{r}
dat %>%
  filter(name_country == namec) %>%
  ggplot(aes(x = year, y = mcpr)) +
  ggplot2::geom_ribbon(data = est, 
    aes(ymin = `low`, y = median, ymax = `up`), fill  = "blue", alpha = 0.1) +
 geom_line(data = est, aes(x = year, y = median), color = "blue") +
  ggplot2::geom_ribbon(data = est2, 
    aes(ymin = `low`, y = median, ymax = `up`), fill  = "red", alpha = 0.1) +
 geom_line(data = est2, aes(x = year, y = median), color = "red") +
  theme_minimal() +
   geom_point() +
  labs(title = namec) + # increase font sizes
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 14))
```


