---
title: "Applied Bayesian Modeling - hierarchical modeling"
author: "Leontine Alkema"
date: "September 5, 2024"
output:
  html_document:
    df_print: paged
  pdf_document:
    number_sections: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(brms)
library(posterior)
library(tidybayes)
```

# Read in radon data 


```{r}
dat0 <- read_csv(here::here("data/radon.csv"))
```


relating the data set to the analysis we want to do: 

- $y_i$ is log(activity)
- county gives county name (fips gives the unique county ID)
- $x_i$ is floor
- $u_i$ is log_ur = log(ura_county)

(the last two are added for the module in which we include predictors)

```{r}
dat <- 
  dat0 %>%
  mutate(y = log(activity), log_ur = log(ura_county))
head(dat)
```

Create summary data set with info for each county:
```{r}
# to plot observations and county means ~ sample sizes, 
# easier to see if sample sizes are slighly jittered
set.seed(12345)

datcounty <- dat %>%
  group_by(fips) %>%
  summarize(nhouses = n(), ybar = mean(y), county = county[1], log_ur = log_ur[1]) %>%
  mutate(nhouses_jitter = nhouses*exp(runif (length(nhouses), -.1, .1)))
ngroups <- dim(datcounty)[1]
head(datcounty)
```

and calculate the overall mean:
```{r}
ybarbar <- mean(dat$y) # population (here state) mean
```

Now let's plot:
```{r}
datcounty %>% 
  ggplot(aes(x = nhouses_jitter, y = ybar)) +
  geom_point() + 
  geom_hline(mapping = aes(yintercept = ybarbar)) + 
  theme_bw()
```

# Model fitting w/o predictors

```{r,  cache = TRUE}
fit <- brm(y ~ (1|county), 
        data = dat, 
        iter = 1000,
        chains = 4,
        cores = getOption("mc.cores", 4))
```

Summary of model fit: 
```{r}
summary(fit)
```

## Visualizing the group-level mean parameters 

Coefficients can be obtained using coef(fit), you can get the help file here:
```{r}
#?coef.brmsfit 
```

Just showing some function calls here first, ie for  mu_alpha:
```{r}
fixef(fit)
```


eta = alpha - mu_alpha (as compared to notation in slides), labeled here as random effects 
```{r}
ranef(fit)$county[,,"Intercept"]
```

saving it in a tibble
```{r}
eta <- as_tibble(ranef(fit)$county[,,"Intercept"], rownames = "county")
head(eta)
```


To get the alpha = eta + mu_alpha, we can use the following call
```{r}
coef(fit, summary = T)$county 
```

again, saving this in a tibble too
```{r}
alphas <-
  coef(fit, summary = T)$county %>% 
  as_tibble(rownames = "county") %>%
  rename(alph = Estimate.Intercept)
alphas 
```

Make the plot of alpha ~ ybar
```{r}
alphas %>%
  left_join(datcounty) %>%
  ggplot(aes(y = alph, x = ybar, size = nhouses)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)
```

Plot of alpha - ybar
```{r}
alphas %>%
  left_join(datcounty) %>%
  ggplot(aes(y = alph - ybar, x = nhouses)) +
  geom_point() +
  geom_hline(yintercept = 0)
```

## Predicting radon levels in unsampled houses and unsampled counties


### Do-it-yourself sampling based approach

We first obtain posterior samples of the hyperparameters and then sample from the predictive distribution for our outcomes of interest, e.g. new observations or group means for new groups.

Extract the posterior samples from the brm-fit: 
```{r}
samp <- as_draws_df(fit)
#dim(samp)
#names(samp)[1:3]
sigmay_s <- samp$sigma
mualpha_s <- samp$b_Intercept
sigmaalpha_s <- samp$sd_county__Intercept
S <- length(sigmay_s)
county_names <- rownames(ranef(fit)$county)
```

Sampling for a new observation, for example from 1st county Aitkin: 
```{r}
county_names[1] 
alpha1_s <- mualpha_s + samp$`r_county[AITKIN,Intercept]`

set.seed(1234) # to make the sampling reproducible
ytilde_s <- rnorm(S, alpha1_s, sigmay_s)
```


Point estimates and 95% CI (using point_interval function from tidybayes package here):
```{r}
point_interval(ytilde_s, .point = mean) 
```

Visualize densities
```{r}
p <- as_tibble(alpha1_s) %>%
  ggplot(aes(alpha1_s, after_stat(density))) +
  geom_histogram(alpha = .5, fill = "blue", bins = 60) +
  theme_minimal() +
  xlab("Log radon") +
  geom_vline(xintercept = mean(alpha1_s), col = "blue")
p +
  geom_histogram(as_tibble(ytilde_s), , bins = 30, mapping = aes(ytilde_s, after_stat(density)), 
               alpha = .5, fill = "red",  linewidth = 1.5) +
  geom_vline(xintercept = mean(ytilde_s), col = "red", linetype = "dashed")
    

```

Sampling for a new group mean: 
```{r}
set.seed(1234) # to make the sampling reproducible
alphanew_s <- rnorm(length(sigmaalpha_s), mualpha_s, sigmaalpha_s)
```

Visualizing the densities
```{r}
p <- as_tibble(mualpha_s) %>%
  ggplot(aes(mualpha_s, after_stat(density))) +
  geom_histogram(alpha = .5, bins = 60) +
  theme_minimal() +
  xlab("Log radon") +
  geom_vline(xintercept = mean(mualpha_s))
p +
  geom_histogram(aes(alphanew_s, after_stat(density)), alpha = .5, fill = "blue", bins = 30, fill = "blue") +
  geom_vline(xintercept = mean(alphanew_s), col = "blue", linetype = "dashed")
    

```


### Can brm-functions do this for me? 

Yes, they can! And when you've fully understood what you're doing, I recommend you take this approach :)  

Prediction for a new house in county 1, we just need a data frame with the county name here (given no predictors)
```{r}
newdata1 <- data.frame(
  county = county_names[1] 
)
ytilde_brm_s <- posterior_predict(fit, newdata = newdata1)

```

Compare the two intervals, should be approximately the same
```{r}
point_interval(ytilde_s, .point = mean) 
point_interval(ytilde_brm_s, .point = mean) 
```


# Model fitting with predictors 

Add group-level predictor floor:
```{r,  cache = TRUE}
fit2 <- brm(y ~ (1+floor|county) + floor,
            data = dat, sample_prior = T, chains = 4,
            iter = 2000, thin = 1,
            cores = getOption("mc.cores", 4))
```


```{r}
summary(fit2)
```

Visualize the fitted regression line (although here just two values for the covariate x, 0 and 1), for each county

```{r}
coefs <- coef(fit2)$county[, 'Estimate', c("Intercept", "floor")]

coefs_tibble <- as_tibble(rownames =  "county", coefs) %>%
  rename(slope = floor)

dat %>%
  full_join(coefs_tibble, by = "county") %>%
  filter(county %in% coefs_tibble$county[1:9]) %>% # just select 4 counties
  ggplot(aes(x = floor, y = y)) +
  geom_point() +
  geom_abline(aes(intercept = Intercept, slope = slope)) +
  geom_abline(aes(intercept = fixef(fit2)[, "Estimate"][1], 
                  slope = fixef(fit2)[, "Estimate"][2]), col = "red") +
  facet_wrap( ~ county)
```





Fit the full model, including county-level uranium 

```{r,  cache = TRUE}
fit3 <- brm(y ~ (1 + floor|county) + log_ur*floor, family = gaussian(), 
            data = dat, sample_prior = T, chains = 4,
            iter = 2000, thin = 1,
            cores = getOption("mc.cores", 4))
```


```{r}
summary(fit3)
```



