---
title: "Radon data"
author: "Leontine Alkema"
date: "September 5, 2024"
output: html_notebook
---

I uploaded in the repo a data set called "radon.csv" that contains the radon data used in course material. The data were obtained through the links indicated in the code below, with some processing steps. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# data

house level data
```{r}
d0 <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat"), 
                header=T, sep=",")
```

county level data
```{r}
cty <- read.table(url("http://www.stat.columbia.edu/~gelman/arm/examples/radon/cty.dat"), 
                  header = T, sep = ",")
```



# processing 

The variables we want
- activity is response variable of interest
- county gives county name (fips gives the unique county ID)
- floor
- ura_county = county level uranium 


process county data:
```{r}
cty <- 
  cty %>% 
  mutate(fips = 1000 * stfips + ctfips) %>% 
  dplyr::select(fips, Uppm) %>%
  rename(ura_county = (Uppm))
```

select MN, deal with zeroes, create fips, select variables, and combine with county data
```{r}
dat <- d0 %>% 
  filter(state=="MN") %>% # Minnesota data only
  mutate(activity = ifelse(activity==0, 0.1, activity)) %>% 
  mutate(fips = stfips * 1000 + cntyfips) %>%   
  dplyr::select(fips, county, floor, activity) %>%
  left_join(cty) 
```

```{r}
dat
```

```{r}
write_csv(dat, here::here("data/radon.csv"))
```





